---
title: "Assessing the Global Divide"
subtitle: "Repository access from the global north and south, before and after COVID"
author: 
  - Jon Wheeler
  - Ngoc-Minh Pham
date: "2023-06-13"
output:
  #powerpoint_presentation: default
    #reference_doc: ref_doc.pptx
  #slidy_presentation: default
  beamer_presentation:
    theme: "PaloAlto"
    colortheme: "crane"
    fonttheme: "structurebold"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../presentations", output_format = "all") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r read data and create plot objects, include=FALSE, message=FALSE}
library(WDI)
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
library(zoo)
library(tidyr)

## Read the aggregated data output by 'aggregate_raw_RAMP_data.py.'
ramp <- read_csv('../raw_data/ramp_sample_2019-2021.csv')

## Import the global region code lookup table.
globalns <- read_csv('../supplemental_data/north_south.csv')

# Convert country codes to lowercase.
# They are lowercase in the RAMP data, so this allows us to merge.
globalns$Countryabbre <- tolower(globalns$Countryabbre)

## Merge the datasets.
## Some rows will be dropped due to the use of 'unknown region' codes
## xkk and zzz in the RAMP data that cannot be joined to the regional
## country coded lookup table.
ramp <- merge(ramp, globalns,
                          by.x = "country", by.y = "Countryabbre") %>% 
  rename(region = Location, iso3c = country) %>% 
  mutate(iso3c = toupper(iso3c), year = year(date))

## For our initial overview, we use RAMP data by itself to demonstrate 
## trends by device type during COVID.

## Get the number of IR in the sample
## Data were cleaned and aggregated using Python
## The sample includes only IR that were in RAMP
## as of Jan 1, 2019
ir_sample <- unique(ramp$repository_id) # 45

## repository_id is otherwise unused
## remove this column
ramp <- ramp %>% 
  select(-repository_id)

## convert region to a factor with labels
## 1 == Global South, 2 == Global North
ramp$region <- as.factor(ramp$region)
levels(ramp$region)[1] <- 'Global South'
levels(ramp$region)[2] <- 'Global North'

# Plot total clicks per day
# with moving average at week (7), month (30), quarter (90)
total_clicks_per_day <- ramp %>% 
  select(date, clicks) %>% 
  group_by(date) %>% 
  summarise(total_clicks = sum(clicks)) %>% 
  mutate(avg_clicks_week = zoo::rollmean(total_clicks, k = 7, fill = NA),
         avg_clicks_month = zoo::rollmean(total_clicks, k = 30, fill = NA),
         avg_clicks_quarter = zoo::rollmean(total_clicks, k = 90, fill = NA)
  ) %>% 
  pivot_longer(names_to = "date_range",
               values_to = "sum_clicks",
               cols = c(total_clicks,
                        avg_clicks_week,
                        avg_clicks_month,
                        avg_clicks_quarter)) %>% 
  ggplot(aes(x = date, y = sum_clicks, color = date_range)) +
  geom_line(mapping = aes(linewidth = date_range,
                          alpha = date_range)) +
  scale_color_manual(breaks = c("avg_clicks_week",
                                "avg_clicks_month",
                                "avg_clicks_quarter"),
                     labels = c("7 day", "30 day", "90 day"),
                     values = c("firebrick1", "blue4", "yellow4", "black")) +
  scale_linewidth_manual(values = c(1.2, 1.2, 1.2, .5),
                         guide = FALSE) +
  scale_alpha_manual(values = c(.7, .7, .7, .4),
                     guide = FALSE) +
  labs(title = "Total clicks per day (2019-2021)",
       subtitle = "RAMP participating institutional repositories",
       x = element_blank(),
       y = element_blank(),
       color = "Rolling averages") +
  theme_linedraw()

## Going to use the 30 day average throughout
## 30 day is smooth but not too smooth
## Total clicks per day, 30 day rolling average
## faceted by global region
total_clicks_30day_ma_region <- ramp %>% 
  select(date, clicks, region) %>% 
  group_by(region, date) %>% 
  summarise(total_clicks = sum(clicks)) %>% 
  mutate(avg_clicks_month = zoo::rollmean(total_clicks, k = 30, fill = NA)) %>% 
  pivot_longer(names_to = "date_range",
               values_to = "sum_clicks",
               cols = c(total_clicks, avg_clicks_month)) %>% 
  ggplot(aes(x = date, y = sum_clicks, color = date_range)) +
  geom_line(mapping = aes(linewidth = date_range,
                          alpha = date_range)) +
  scale_color_manual(breaks = c("avg_clicks_month"),
                     labels = c("30 day"),
                     values = c("red", "black")) +
  scale_linewidth_manual(values = c(1, .5),
                         guide = FALSE) +
  scale_alpha_manual(values = c(.9, .4),
                     guide = FALSE) +
  facet_wrap(facets = vars(region)) +
  labs(title = "Total clicks per day (2019-2021)",
       subtitle = "By global region",
       x = element_blank(),
       y = element_blank(),
       color = "Rolling average") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


## Above, filtered to just 2020
total_clicks_30day_ma_region_2020 <- ramp %>% 
  select(date, clicks, region, year) %>% 
  group_by(year, region, date) %>% 
  filter(year == "2020") %>% 
  summarise(total_clicks = sum(clicks)) %>% 
  mutate(avg_clicks_month = zoo::rollmean(total_clicks, k = 30, fill = NA)) %>% 
  pivot_longer(names_to = "date_range",
               values_to = "sum_clicks",
               cols = c(total_clicks, avg_clicks_month)) %>% 
  ggplot(aes(x = date, y = sum_clicks, color = date_range)) +
  geom_line(mapping = aes(linewidth = date_range,
                          alpha = date_range)) +
  scale_color_manual(breaks = c("avg_clicks_month"),
                     labels = c("30 day"),
                     values = c("red", "black")) +
  scale_linewidth_manual(values = c(1, .5),
                         guide = FALSE) +
  scale_alpha_manual(values = c(.9, .4),
                     guide = FALSE) +
  facet_wrap(facets = vars(region)) +
  labs(title = "Total clicks per day (2020)",
       subtitle = "By global region",
       x = element_blank(),
       y = element_blank(),
       color = "Rolling average") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


## The full 2019-2021 time series
## clicks by device
## faceted by region
clicks_device_region_all <- ramp %>% 
  select(date, clicks, region, device) %>% 
    group_by(region, device, date) %>% 
    summarise(total_clicks = sum(clicks)) %>% 
    mutate(avg_clicks_month = zoo::rollmean(total_clicks, k = 30, fill = NA)) %>% 
    pivot_longer(names_to = "date_range",
                 values_to = "sum_clicks",
                 cols = c(total_clicks, avg_clicks_month)) %>% 
    ggplot(aes(x = date, y = sum_clicks, color = date_range)) +
    geom_line(mapping = aes(linewidth = date_range,
                            alpha = date_range)) +
    scale_color_manual(breaks = c("avg_clicks_month"),
                       labels = c("30 day"),
                       values = c("red", "black")) +
    scale_linewidth_manual(values = c(.75, .5),
                           guide = FALSE) +
    scale_alpha_manual(values = c(.8, .6),
                       guide = FALSE) +
    facet_wrap(facets = vars(region, device)) +
    labs(title = "Total clicks per day (2019-2021)",
         subtitle = "By global region and device",
         x = element_blank(),
         y = element_blank(),
         color = "Rolling average") +
    theme_linedraw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))



## As above, filtered to 2020
## faceted by device and region
clicks_device_region_2020 <- ramp %>% 
  select(date, clicks, region, device, year) %>% 
  group_by(year, region, device, date) %>% 
  filter(year == "2020") %>% 
  summarise(total_clicks = sum(clicks)) %>% 
  mutate(avg_clicks_month = zoo::rollmean(total_clicks, k = 30, fill = NA)) %>% 
  pivot_longer(names_to = "date_range",
               values_to = "sum_clicks",
               cols = c(total_clicks, avg_clicks_month)) %>% 
  ggplot(aes(x = date, y = sum_clicks, color = date_range)) +
  geom_line(mapping = aes(linewidth = date_range,
                          alpha = date_range)) +
  scale_color_manual(breaks = c("avg_clicks_month"),
                     labels = c("30 day"),
                     values = c("red", "black")) +
  scale_linewidth_manual(values = c(.75, .5),
                         guide = FALSE) +
  scale_alpha_manual(values = c(.8, .6),
                     guide = FALSE) +
  facet_wrap(facets = vars(region, device)) +
  labs(title = "Total clicks per day (2020)",
       subtitle = "By global region and device",
       x = element_blank(),
       y = element_blank(),
       color = "Rolling average") +
  theme_linedraw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

## A previous analysis on a smaller dataset
## indicated that users in the global south
## are more likely to use mobile devices
## than users from the global north.
## This is evident from above, but
## we now make this comparison with the larger
## dataset and augment with World Bank population data
# Get World Bank population data

# Web search gives the following as total population indicator
# https://data.worldbank.org/indicator/SP.POP.TOTL
world_pop <- WDI(indicator = c(population = 'SP.POP.TOTL'),
                 start = 2019,
                 end = 2021)

# need to add region data to population data
global_region <- read_csv('../supplemental_data/north_south.csv')

# merge region data to population data
# drop columns we won't use
pop_region <- merge(world_pop, global_region, 
                    by.x = "iso3c",
                    by.y = "Countryabbre") %>% 
  select(country, iso3c, year, population,
         Location) %>% 
  rename(region = Location)

# make Location a factor
pop_region$region <- as.factor(pop_region$region)
levels(pop_region$region)[1] <- 'Global South'
levels(pop_region$region)[2] <- 'Global North'

# join RAMP data with population data
ramp_pop <- inner_join(ramp, world_pop,
                       by = c("iso3c", "year"))

# We can look at clicks by device and by region
clicks_by_device_region <- ramp_pop %>% 
  select(region, population, device, clicks, year) %>% 
  group_by(region, device, population, year) %>% 
  summarise(clicks = sum(clicks)) %>% 
  group_by(region, device, year) %>% 
  summarize(pop = sum(population),
            clicks = sum(clicks)) %>% 
  mutate(clicks_weighted = (clicks/pop)*100) %>%
  arrange(desc(region))

# Note we are using population data to allow us
# to derive population weighted click values
# First, plot using unweighted values
options(scipen = 999)
unweighted_clicks_device <- clicks_by_device_region %>% 
  ggplot(aes(x = year, y = clicks, color = region)) +
  geom_line(linewidth = 1) +
  facet_wrap(facets = vars(device)) +
  labs(title = "Clicks by device (2019-2021)",
       x = element_blank(),
       y = element_blank(),
       color = "Region") +
  theme_linedraw() +
  theme(axis.text.x = element_blank())


# Now use clicks weighted by population
weighted_clicks_device <- clicks_by_device_region %>% 
  ggplot(aes(x = year, y = clicks_weighted, color = region)) +
  geom_line(linewidth = 1) +
  facet_wrap(facets = vars(device)) +
  labs(title = "Clicks by device (2019-2021)",
       subtitle = "Weighted by population (per 100 people)",
       x = element_blank(),
       y = element_blank(),
       color = "Region") +
  theme_linedraw() +
  theme(axis.text.x = element_blank())
```

# Background

## About the Repository Analytics and Metrics Portal (RAMP)

- Aggregate search engine performance metrics
  - Google search properties
  - Consistent reporting across repositories and platforms
  - Longitudinal data (since 2017 for some IR)
  - No-cost, online dashboards for participating repositories
- Open datasets
  - [Subset of 35 IR, Jan 1 - May 31, 2019](https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.fbg79cnr0)
  - [All participating IR (with consent), 2017 - 2021](https://datadryad.org/search?q=repository+analytics+and+metrics+portal)
  - [Complete documentation is available.](https://digitalrepository.unm.edu/ulls_fsp/141/)

## Objectives

Invite discussion about future research and the future of RAMP.

- Previous analysis of the Jan - May 2019 subset (Spring 2020: Pre-COVID!)
  - Access of IR content is unevenly distributed across IR and items
  - Users from the global south are far more likely to use mobile devices
  - Weak but significant correlation between population and access of IR content
- Redo previous analysis with more data
  - _**Identify**_ COVID impact, if any, on access of IR content
  - Surface trends in RAMP data and identify external data source of research 
  interest
  - Define research questions and invite further analysis and participation

## About the present sample

- Three years of data for 45 repositories
  - Repositories not in RAMP before January 1, 2019 excluded
  - Date range January 1, 2019 - December 31, 2021
- Limitations
  - Cannot assess or compare with longer term trends
  - Sample is not representative
    - Global regions
    - Languages
  - Google Search Console API is the sole data source

# Trends

## General seasonal trend (2019-2021)

Daily totals with 7 day, 30 day, and 90 day moving averages.

```{r seasonal trends across dataset, message=FALSE}
total_clicks_per_day
```

## COVID disruption of seasonal trend (2020)

Daily totals with 30-day moving average.

```{r COVID disruption of trend 2020}
total_clicks_30day_ma_region_2020
```

## COVID 2020 disruption by device

Daily totals with 30-day moving average.

```{r COVID 2020 disruption by device}
clicks_device_region_2020
```

## Overall clicks by device

```{r}
unweighted_clicks_device
```

## Clicks by device, weighted by popuation (World Bank)

```{r}
weighted_clicks_device
```

## Data Sources (other than RAMP data)

- International Organization for Standardization. (2020),
“ISO 3166 Country Codes”, ISO, Nongovernmental organization, 
available at: https://www.iso.org/iso-3166-countrycodes.html 
(accessed 12 April 2020).
- Meta contributors. (2020), “List of countries by regional classification”, 
Wikimedia Meta-Wiki, Meta, discussion about Wikimedia projects, 1 April, 
available at:
https://meta.wikimedia.org/w/index.php?title=List_of_countries_by_regional_classification&oldid=19943813 (accessed 12 April 2020).
- World Bank. (2023), "Population, total," The World Bank Group, https://data.worldbank.org/indicator/SP.POP.TOTL.